---
title: "Divvy Bike Share 2024 Analysis"
author: "Dmytro Khavrel"
output: 
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 2
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(scales)
```

# 🚲 Introduction: About Divvy

Divvy is Chicago’s official bike share system, launched in 2013 and operated by **Lyft** in partnership with the **Chicago Department of Transportation (CDOT)**. It has grown into one of the largest bike share programs in North America, with:

- Over **800 docking stations** across Chicago and nearby suburbs,
- A fleet of **traditional pedal bikes** and **electric Divvy eBikes**,
- Options for **casual riders** (pay-as-you-go) and **annual members**.

Divvy continues expanding its network and electrified fleet, while aiming to convert more casual users into long-term members. This analysis uses full-year trip data from 2024 to explore ridership trends and support Divvy’s business goals.

## 🎯 Project Goals

This project answers key business questions:

1. **How do annual members and casual riders use Divvy bikes differently?**
2. **Why would casual riders buy annual memberships?**
3. **How can Divvy use digital media to influence casual riders to become members?**

We will use R to process, clean, analyze, and visualize real trip data, leading to insights and marketing recommendations.

# 📥 Load & Prepare Data

```{r}
#setwd("Divvy Trips 2024")
months <- sprintf("%02d", 1:12)
files <- paste0("2024", months, "-divvy-tripdata.csv")

divvy_2024 <- files %>% map_df(read_csv)

# Ride duration
divvy_2024$ride_length <- difftime(divvy_2024$ended_at, divvy_2024$started_at)
divvy_2024$ride_length <- as.numeric(divvy_2024$ride_length)
```

```{r}
divvy_2024_clean <- divvy_2024 %>%
  distinct(ride_id, .keep_all = TRUE) %>%
  drop_na(ride_id, started_at, ended_at, member_casual, rideable_type,
          start_station_name, end_station_name, start_lat, start_lng, end_lat, end_lng) %>%
  mutate(across(c(start_station_name, end_station_name), str_trim)) %>%
  filter(
    ride_length > 0,
    start_station_name != "HQ QR",
    between(start_lat, 40.73, 42.67),
    between(end_lat, 40.73, 42.67),
    between(start_lng, -88.94, -86.93),
    between(end_lng, -88.94, -86.93),
    nchar(start_station_name) > 1,
    nchar(end_station_name) > 1
  )

# Add day of the week
divvy_2024_clean <- divvy_2024_clean %>%
  mutate(date = as.Date(started_at),
         day_of_week = format(date, "%A"))
```

# 📊 Descriptive Statistics

```{r}
summary(divvy_2024_clean$ride_length)
```

## 🚲 Average Ride Length by Rider Type

```{r}
divvy_2024_clean %>%
  group_by(member_casual) %>%
  summarise(
    mean_duration = mean(ride_length),
    median_duration = median(ride_length),
    max_duration = max(ride_length),
    min_duration = min(ride_length)
  )
```

# 📈 Ride Patterns

## 🗓️ Weekly Ridership Count

```{r}
divvy_2024_clean %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), .groups = "drop") %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Number of Rides by Rider Type", y = "Ride Count", x = "Weekday") +
  theme_minimal()
```

## ⏱️ Average Duration per Day

```{r}
divvy_2024_clean %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(average_duration = mean(ride_length), .groups = "drop") %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Average Ride Duration by Rider Type", y = "Duration (sec)", x = "Weekday") +
  theme_minimal()
```

## 📍 Ride Start Heat Map

```{r}
divvy_2024_clean %>%
  ggplot(aes(x = start_lng, y = start_lat)) +
  geom_point(alpha = 0.05, color = "grey30") +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon", color = NA, contour = TRUE) +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  coord_quickmap() +
  facet_wrap(~ member_casual) +
  labs(
    title = "Ride Start Location Heat Map by User Type",
    x = "Longitude",
    y = "Latitude",
    fill = "Density"
  ) +
  theme_minimal()
```

## 📍 Ride End Heat Map

```{r}
divvy_2024_clean %>%
  ggplot(aes(x = end_lng, y = end_lat)) +
  geom_point(alpha = 0.05, color = "grey30") +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon", color = NA, contour = TRUE) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  coord_quickmap() +
  facet_wrap(~ member_casual) +
  labs(
    title = "Ride End Location Heat Map by User Type",
    x = "Longitude",
    y = "Latitude",
    fill = "Density"
  ) +
  theme_minimal()
```


# 📤 Export Clean Data

```{r}
write.csv(divvy_2024_clean, "divvy_2024_member_vs_casual.csv", row.names = FALSE)
```

# 🔍 Key Insights

## ✅ How do annual members and casual riders differ?

Casual riders tend to:

* Ride more on weekends.
* Take longer trips.
* Use bikes for recreation.

Annual members are more likely to:

* Ride during weekdays.
* Take shorter trips.
* Use bikes for commuting or errands.

## 💡 Why might casual riders convert to membership?

* Regular weekend riders could save money by switching to an annual plan.
* Convenience: easier access, no per-ride hassle.
* Better value for frequent short rides.

## 📣 Digital Marketing Strategy Suggestions

* Targeted ads on weekends showing savings based on their usage.
* Promote features like unlock speed and exclusive stations.
* Offer trial discounts during peak casual riding months (e.g., summer).

## 📌 Conclusion

Divvy can grow its annual membership base by:

* Understanding behavioral patterns,
* Personalizing offers based on usage,
* Making value propositions clear through data-driven digital media.